Datelife Reports

### 2020-04-30 paper_3-main.Rmd
ejm comments

Summarizing high-quality chronograms might also imply summarizing evolutionary hypothesis. This could be enlightening from certain point of view, since it could help getting a single global evolutionary history for a lineage. It could also be really misleading, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in some chronograms and not from the summary chronogram.
Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart]. This will be implemented in future `datelife` versions.

<!-- still requires more research. -->
<!-- A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced. However, this is just an assumption.  -->

Alternatively, one could try to choose the "best" chronogram. Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms. Some characteristics that are often cited as a measure of improved quality in dating studies addressing long-studied organisms are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals. To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the near future.

@@

In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance. Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ [@Webb2008]) for several taxa lacking this information. Making up branch lengths in this or other ways is accepted in scientific publications: @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data and trees that are reconstructed using taxonomy.

@@

Divergence time information is key to many areas of evolutionary studies: trait evolution,
diversification, biogeography, macroecology and more. Generating this information is difficult,
especially for those who want to use phylogenies but who are not systematists, or
do not have the time to acquire and develop the necessary knowledge and data curation skills
to produce chronograms _de novo_. Knowledge on clade ages is also crucial for
science communication and education.

<!--Mention again uses shown here and outline potential uses not shown here-->
`datelife` allows an easy and fast obtention, as well as comparison of publicly available information
on time of lineage divergence, providing a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, allowing identification of regions that require more research or that have conflicting information.
<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
Both summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research.  `datelife` helps with awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

>>>>

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses. For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations. Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence). Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms. Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not-->
However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis. Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram. <!-- Example -->

<!-- still requires more research. -->
<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced. However, this is just an assumption.  -->

Alternatively, one could try to choose the "best" chronogram from a set of possible evolutionary hypotheses. Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms. Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals. To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.

@@

In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008]. Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option). Making up branch lengths in this or other ways is accepted in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data adn the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.

@@

Divergence time information is key to many areas of evolutionary studies: trait evolution,
diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms _de novo_ is difficult,
especially for those who want to use phylogenies but who are not systematists, or
do not have the time to acquire and develop the necessary knowledge and data curation skills. Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.

<!--Mention again uses shown here and outline potential uses not shown here-->
`datelife` allows easy and fast summarization of publicly available information
on time of lineage divergence. This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.  `datelife` helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
